---
title: "FinalFigures_160901"
author: "Bengt Sennblad"
output: html_document
date: "`r format(Sys.time(), '%d %B, %Y')`"
editor_options: 
  chunk_output_type: console
---

```{r global_options, include=FALSE}
knitr::opts_chunk$set(echo=FALSE, warning=FALSE, message=FALSE,eval=TRUE)
```

First load required libraries and some convenient helper functions; set directory for reading data

```{r libs, results="hide", eval=TRUE}
source(file="Helper.R", chdir=T,echo=FALSE, verbose=FALSE)
libs=c("abind","ggplot2","gridExtra","R.utils","plyr","knitr","WriteXLS", "tibble","gridGraphics") # NB! I have noted that gridGraphics might require (re)installation of ImageMagick:( (issue was on Mac after install of MacosX 10.13
checklibs(libs)

in_dir="R-files/"
out_dir="FinalOutput_181018/"
## To recreate old (pre 160901) results using an erroneous rotation matrix for ii-vi, set the following variable to TRUE
showOldErrorneousVersion = FALSE
if(showOldErrorneousVersion == TRUE){
 out_dir="FinalOutput/"
}
dir.create(out_dir, showWarnings=FALSE)

#Default scatterpot point size 
myscatterpointsize=0.5
mybasesize = 5
PAM50cols=c("Basal"="#E31A1C", "Her2"="#FB9A99","LumA"="#1F78B4","LumB"="#A6CEE3","Normal"="#33A02C")

# Standard scatterplot of two variables
scatterplot<-function(data,x,y, hadjust=-0.2){
  ret=ggplot(data,aes_string(x=x, y=y))+
    stat_bin2d(bins=75)+
    scale_fill_gradientn(limits=c(0,150), breaks=seq(0, 150, by=25), colours=rainbow(4))+
    theme_classic(base_size=8)+
    theme(axis.line.y = element_line(colour = "black", linetype="solid", size=0.5),axis.line.x = element_line(colour = "black", linetype="solid", size=0.5))+
    theme(plot.title = element_text(hjust=hadjust,margin = margin(l=2,r=2,t = 0, b = -5)))+
    theme(legend.justification=c(0,1),legend.position=c(0.9,0.35),legend.title=element_blank(), legend.key.size=unit(7, "pt"),legend.text=element_text(size=4))
  return(ret)
}


# saturation transformation function
satf<-function(y){
  y=(y+1)/2
  require(boot)
  return(logit(y))
}

mkprop<-function(y){
  return(y=(y+1)/2)
}

# This is mainly a helper function called by other plot functions
# returns a list with a data.frame with a saturation transformed 
# version of y and a function for the regression line between the 
# transformed y and x
satdata<-function(data, x,y,satname){
  # Regression between rotated mRNA and correlations
  #regcorrxp=lm(satf(data[,y])~data[,x])
  data$ytr=mkprop(data[,y])
  f=as.formula(paste("ytr","~", paste0('`',x, '`',collapse="+")))
  regcorrxp=glm(f, data=data, family=quasibinomial(link="logit"), na.action="na.omit")
  data$residuals=regcorrxp$residuals
  pars=regcorrxp$coefficients
  k=pars[2]
  m=pars[1]
  regline=function(x){
    return(k*x+m)
  }
  ret <- data
  ret[,satname] <- satf(ret[,y])
  return(list(ret,regcorrxp)) 
}

# plots a scatterplot of x and saturation-transformed y, with density 
# of plots indicated by stat_bin2d, and a regression line
binlogitplot<-function(data, x,y,cov){
  # get the transformed data and regression line function
  tmpname=paste("transformed",y,sep="_")
  tmp = satdata(data,cov,y,tmpname)
  # the plot
  regline=function(x){
    pars=tmp[[2]]$coefficients
    k=pars[2]
    m=pars[1]
    return(k*x+m)
  }

  ret = ggplot(tmp[[1]],aes_string(x=x, y=tmpname))+
    theme_classic(base_size=8)+
    theme(axis.line.y=element_line(colour="black",linetype="solid",size=0.5),axis.line.x = element_line(colour = "black", linetype="solid", size=0.5))+
    theme(plot.title = element_text(hjust=-0.2,margin = margin(l=2,r=2,t = 0, b = -5)))+
    stat_bin2d(bins=75,)+ 
    scale_fill_gradientn(limits=c(0,150),breaks=seq(0,150,by=25),colours=rainbow(4))+
    theme(legend.justification=c(1,0),legend.position=c(1.0,0.1),legend.title=element_blank(), legend.key.size=unit(7, "pt"),legend.text=element_text(size=4))+
    stat_function(fun=regline)
#    stat_function(fun=tmp[[2]])
  return(ret)    
}



# plots a scatterplot of x and logit-transformed y, the subset defined 
# by meta in a different color and a regression line based on logit(y)~cov
grouplogitplot<-function(data, x, y, cov, meta){
  tmpname = paste("transformed",y,sep = "_")
  tmp = satdata(data,cov, y, tmpname)
  
  regline=function(x){
    pars=tmp[[2]]$coefficients[x]
    k=pars[2]
    m=pars[1]
    return(k*x+m)
  }

  metaset=data[rownames(data) %in% exprdata$metagene[[meta]],]
  tmp2=satdata(metaset,x,y,tmpname)
  ret=ggplot(tmp[[1]],aes_string(x=x, y=tmpname))+
    theme_classic(base_size=8)+
    theme(axis.line.y=element_line(colour="black",linetype="solid",size=0.5),axis.line.x = element_line(colour = "black", linetype="solid", size=0.5))+
    theme(plot.title = element_text(hjust=0.5,margin = margin(l=2,r=2,t = 0, b = -5)))+
    geom_point(colour="red", size=myscatterpointsize)+
    geom_point(data=tmp2[[1]],aes(x=tmp2[[1]][,x], y=tmp2[[1]][,tmpname]),colour="black", size=myscatterpointsize)+
    stat_function(fun=regline)
#    geom_point(data=metaset,aes(x=metaset[,var], y=satf(metaset[,"correlations"])),colour="black", size=myscatterpointsize)+
  return(ret)
  }

# Boxplot of data$vakue classified by group
bxplot<-function(data, group, value){
  ret=ggplot(data, aes(factor(group), value))+
    geom_boxplot()+
    theme_classic(base_size=8)+
    theme(axis.line.y = element_line(colour = "black", linetype="solid", size=0.5),axis.line.x = element_line(colour = "black", linetype="solid", size=0.5))+
    theme(plot.title = element_text(hjust=-0.12,margin = margin(t = 0, b = -5)))+
    theme(axis.text.x = element_text(angle=45,vjust=1,hjust=1))
  return(ret)      
}

tumourplot<-function(data,x,y,fill){
  ret=ggplot(data,aes(x=reorder(data[,x],data[,y]), y=data[,y], fill=data[,fill]))+
    scale_fill_manual(values=PAM50cols)+
    geom_bar(stat="identity")+
    theme_classic(base_size=8)+
    theme(axis.line.y = element_line(colour = "black", linetype="solid", size=0.5),axis.line.x = element_line(colour = "black", linetype="solid", size=0.5))+
    theme(axis.text.x = element_text(angle=90,vjust=1,hjust=1,size=4),text=element_text(size=5))+
    theme(plot.title = element_text(hjust=-0.05,margin = margin(l=2,r=2,t = 0, b = -5)))+
    theme(legend.justification=c(1,0),legend.position=c(0.2,0.8),legend.title=element_blank(), legend.key.size=unit(6, "pt"),legend.text=element_text(size=4))+
    geom_hline(aes(yintercept=mean(data[,y])), linetype="dashed")+
    annotate("text", y=mean(data[,y])+0.02,x=4.5,
           label=paste("mean=",round(mean(data[,y]),3)),size=1.5)
  return(ret)
}


```

Load the R-object containing data (construct it if it does not exist).

```{r loaddata, eval=TRUE}
# Create infile if needed
if(!file.exists(paste(in_dir,"tumourExpressionData", sep="/"))){
  source("formatData.R", chdir=TRUE)
}
# load expression working data as an R object called exprdata
load(paste(in_dir,"tumourExpressionData",sep="/"))
# Identify intersection of gene names and tumour names
genes=intersect(rownames(exprdata$protexpr),rownames(exprdata$rnaexpr))
tumours=intersect(colnames(exprdata$protexpr), colnames(exprdata$rnaexpr))
# create a 3D array of corresponding expression data, with tumours as cols and genes as rows, add names for the new dimension 
data=abind(exprdata$protexpr[genes,tumours],exprdata$rnaexpr[genes,tumours], along=3,new.names=c(NULL,NULL,c("prot","rna")))
```

We first perform the correlation analysis, and focus on Spearman correlations.

```{r makecorr, eval=TRUE, include=FALSE}
method="spearman"
# Create correlations and plots using spearman correlations
corr_label=paste(capitalize(method),"correlations") # label for correlation axis

source(file="Helper.R", echo=FALSE, verbose=FALSE)

# create the correlations over tumours and over genes
corr_tumour = doCorrDiagPairs(data, meth=method, meta=data.frame(PAM50=exprdata$metatumour[,"PAM50"]), indexname="tumour")
corr_genes = doCorrDiagPairs(data,meth=method, marg=c(1), indexname="gene")
#we want to just correlate for each tumours using the 300 best correlations
all = rownames(corr_genes)
best300 = rownames(corr_genes[order(corr_genes$correlations, decreasing=TRUE),])[0:300]
subsets=list(all=all,best300=best300)
tmp=doCorrDiagPairs(data[best300,,],meth=method,indexname="tumour")
names(tmp)= paste("Best300",names(tmp),sep=".")

corr_tumour=merge(corr_tumour,tmp[,c("Best300.correlations","Best300.Pval","Best300.Padjust")],by="row.names")
rownames(corr_tumour) = corr_tumour$Row.names
corr_tumour$Row.names=NULL
# corr_tumour[["Best300.correlations"]] = tmp$correlations
# corr_tumour[["Best300.Pval"]] = tmp$Pval
# corr_tumour[["Best300.Padjust"]] = tmp$Padjust


# corr_tumour[["Best300"]]=doCorrDiagPairs(data[best300,,],meth=method,meta=data.frame(PAM50=exprdata$metatumour[,"PAM50"]),indexname="tumour")
```


We save the the correlations as tab-delimited text files ( to make it available to users.

```{r savecorr}
outfile1=paste(out_dir,paste("Lehtio.all",method,"tumour_correlationdata.txt",sep="."),sep="/")
write.table(corr_tumour,file=outfile1,sep="\t", row.names=FALSE)

outfile2=paste(out_dir,paste("Lehtio.all",method,"gene_correlationdata.txt",sep="."),sep="/")
write.table(corr_genes,file=outfile2,sep="\t", row.names=FALSE)

```
Saved correlations to files `r outfile1` and `r outfile2`


We also want to, eventually, relate correlations to variance statistics. So compute these stats as well.

```{r makestats,eval=TRUE}
# create stats for prot and rna data and combined them with correlations
statp = getStat(data[,,"prot"],c(1))
names(statp)=sub("^","protein_",names(statp))
statr = getStat(data[,,"rna"],c(1))
names(statr)=sub("^","mRNA_",names(statr))
statcorr = cbind(statp,statr,corr_genes)
```
and save correlations and stats in common data.frame corrxp
```{r,merge}
corrxp=corr_genes
for(stat in c("sd")){
  myscatterpointsize=1
  rvar=paste("mRNA", stat, sep="_")
  pvar=paste("protein", stat, sep="_")
  sds=data.frame(statcorr[[rvar]],statcorr[[pvar]],row.names=rownames(statcorr))
  colnames(sds)=c(rvar,pvar)
  sds$gene=rownames(sds)
  
  # Create a dataframe with rotated sds and correlations
  corrxp = join(sds,corrxp, by="gene", type="inner")
  rownames(corrxp) = corrxp$gene
}
```

# Modeling Protein-mRNA correlation dependence on mRNA and protein standard deviation
To investigate how well the standard deviation of mRNA and protein expression explains the correlation between protein and mRNA expression, we first look at the $sd_{protein}$ and $sd_{RNA}$ separately in regression analyses. However, since correlation coefficients are bounded by in the interval (-1,1), linear regression will not work well. In fact, this is simlar to the case when modeling proportions -- these are bounded betweein the interval (0,1) and often shows a sigmoidal ('S'-shaped) distribution when plotted to the independent variable.

A standard solution is to use a generalized linear model (glm) with a logit-transformation of the correaltions and a binomial distribution modeling the residuals, $logit(y) = \alpha + \beta x + \epsilon$, where $\alpha$ is the intercept, $\beta$ is the effect size of $x$ and $\epsilon$ is an error term representing the residuals, i.e., the difference between the observed $logit(y)$ value and that predicted by the regression; in this model $\epsilon$ is modeled to follow a binomial distribution.

We will first transform the correlation coefficient values, $y$, to the interval (0,1) (simply using $y'=(y+1)/2$) and then apply the glm above for regression of correlation coefficients to each mRNA or protein statistic alone (univariate regression: correlations~stat_mRNA or correlations~stat_protein). Figure S11A_D shows the results. 

Subfigs A-B shows the results for mRNA sd, while subfigs C-D show the results for protein sd. A and C shows plain scatter plots of the untransformed correlation as a function of the mRNA or protein sd, respectively We can note the almost (lower part of 'S' is missing) sigmoidal behaviour of the correlations as a function of sd (similar to the standard "proportion" distribution).

Subfigs B and D shows scatterplots of the logit-transformed correlations (transformed to interval (0,1)) as a function of mRNA or protein sd, respectively; the regression line for a univariate glm, as described above, is indicated.

Clearly, both mRNA sd and protein sd are explanatory fro the mRNA-protein correlation. It is therefore reasonable to create a bivariate glm for mRNA-protein correlation as a function of both mRNA sd and protein sd, $logit(y')=\alpha+\beta x + \gamma z + \epsilon$, with $x$ being mRNA sd and $z$ protein sd.

Table S7 shows the regression coefficient and the p-value for that coefficient for the two univariate and the bivariate regressions. There are some overlap in the signals from protein and mRNA sd, as the coefficients are clearly lower in the bivariate regression. However, none goes down to zero, hence they appear also to provide independent signals.

Distribution of selected protein groups in scatterplots of logit-transformed Spearman correlation vs combined mRNA-protein SD is shown in Figure S11E-R.

```{r,prefig1, eval=TRUE, fig.height=6}
 # pl will hold the raw  scatterplots
gl=list()
j=0

pl=list()
i=0
for(type in c("mRNA", "protein")){
  var= paste(type, "sd", sep="_")
  # First untransformd plot
  i=i+1
  pl[[i]]=scatterplot(corrxp,var,"correlations", hadjust=-0.2)+
    scale_x_continuous(sub("sd","SD",var))+
    scale_y_continuous("mRNA-protein correlation") 
  # then transformed plot with regression 
  i=i+1
  pl[[i]]=binlogitplot(corrxp,var,"correlations", cov=c("mRNA_sd","protein_sd"))+
    scale_x_continuous(sub("sd","SD",var))+
    scale_y_continuous("logit-transformed\nmRNA-protein correlation") 
}
```

## Table S3 : regression coefficients

### Legend
Effect sizes and p-values from univariate and bivariate regressions(glm:logit+binomial) using the statistics, mean , sd, and cv

### The table
```{r new2}
tb=data.frame(stat=list())
for(stat in c("sd")){
  rvar= paste("mRNA", stat, sep="_")
  pvar= paste("protein", stat, sep="_")
  for(var in c(rvar, pvar)){
    fit=summary(glm(mkprop(corrxp[,"correlations"])~corrxp[,var], family=quasibinomial(link="logit"), na.action="na.omit"))$coefficients
    tb[var,"univar_effect"] = fit[2,1]
    tb[var,"univar_P-value"] = sprintf("%.4g",fit[2,4])
  }
  # now the bivariate model
  fit=summary(glm(mkprop(corrxp[,"correlations"])~corrxp[,rvar]+corrxp[,pvar], family=quasibinomial(link="logit"), na.action="na.omit"))$coefficients
  i= 2
  for(var in c(rvar, pvar)){
    tb[var,"bivar_effect"] = fit[i,1]
    tb[var,"bivar_P-value"] = sprintf("%.4g",fit[i,4])
    i=i+1
  }
}
kable(tb, align="l",row.names=T) #, caption="Effect sizes and p-values from univariate and bivariate regressions(glm:logit+binomial) using the statistics, mean , sd, and cv")
```

```{r, writeS6table,results="hide"}
outfile = paste(out_dir,"Lehtio_S3_regression_coefficients.XLSX",sep="/")
WriteXLS(tb, ExcelFileName=outfile,row.names=T)
```
Save Table as `r outfile`.





## Supplementary Figure S11

###Legend:
"A-D) mRNA-protein correlation as a function of standard deviation (SD) for mRNA and protein expression values. For each statistics (see x-axis legend), A and B shows, as a function of mRNA SD, a scatterplot of untransformed correlation coefficients and a scatterplot of logit-transformed correlation with a glm-regression line, respectively, and C and D shows the same for protein SD. E-R) Distribution of selected protein groups in scatterplots of logit-transformed Spearman correlation vs mRNA or protein SD. E-F)  Mammaprint (Tian_2010), G-H) PAM50, I-J) Protein targets of_FDA appr. Drugs, K-L) Ribosomal, M-N) Spliceosome, O-P) COSMIC_database_and__Nik-Zainal_93_BC_drivers, Q-R) COSMIC_BC_drivers_and_Nik-Zainal_93_BC_drivers. Black dots indicate the specified protein group, while red dots represent all other proteins.  The black line indicates the marginal regression line for mRNA or protein sd, i.e., logit(y)~b_ix_i+c for i in (1,2), where x_1 is the mRNA SD and x_2 the protein SD, and b_1 and b_2 are the corresponding effect sizes for x_1 and x_2, respectively, estimated from the bivariate regression model logit(y)~b_1 x_1+b_2 x_2 + c."

### The fig
```{r, proteingroups, eval=TRUE}
cvar="mRNA_sd+protein_sd"
myvars=list(c("mRNA_sd","protein_sd"),"mRNA_sd","protein_sd")

priorPosterior=list(
  Predefined=c("Oncotype_DX",
          "Mammaprint_(Tian_2010)",
          "PAM50",
          "Protein_targets_of_FDA_appr._Drugs",
          "Ribosomal",
          "Spliceosome",
          "Transcription_factors",
          "Transporters",
          "COSMIC_database_and__Nik-Zainal_93_BC_drivers",
          "COSMIC_BC_drivers_and_Nik-Zainal_93_BC_drivers"),
  Enriched=c("KEGG_RNA_polymerase",
              "KEGG_Ribosome",
              "KEGG_Oxidative_phosphorylation",
              "KEGG_Spliceosome",
              "KEGG_Tryptophan_metabolism",
              "KEGG_Valine,_leucine_and_isoleucine_degradation",
              "KEGG_Fatty_acid_metabolism",
              "KEGG_Steroid_hormone_biosynthesis",
              "Hallmark_oxidative_phosphorylation",
              "Hallmark_estrogen_response_early",
              "Hallmark_estrogen_response_late",
              "Hallmark_interferon_gamma_response",
              "Hallmark_G2M_checkpoint",
              "Hallmark_interferon_alpha_response",
              "Hallmark_E2F_targets",
              "Hallmark_MTORC1_signaling")
)
pmetas=list() # Store the residual comparison results for both prior and posterior groups
for(p in c("Predefined","Enriched")){
  #pmeta = data.frame(name=names(exprdata$metagene))
  pmeta = data.frame(name=priorPosterior[p])
  colnames(pmeta)=c("name")
  
  rownames(pmeta) = gsub(" ","_", pmeta$name)
  # fill in mean correlation for each meta group
  for(meta in rownames(pmeta)){
    pmeta[meta,"meancorr"]=mean(corrxp$correlations[rownames(corrxp) %in% exprdata$metagene[[meta]]])
  }
  
  
  for(meta in rownames(pmeta)){
    for(var in myvars){
      # var="combined mRNA and protein sd"
      # tmp = satdata(corrxp,var,"correlations", "transformed_correlations")[[1]]
      # regcorrxp=as.data.frame(lm(tmp[,"correlations"]~tmp[,var])$residuals)
      newvar=paste0(var,collapse="+")
      corrxp$ytr=mkprop(corrxp[,"correlations"])
      f=as.formula(paste("ytr","~", paste0('`',var, '`',collapse="+")))
      regcorrxp=as.data.frame(glm(f, data=corrxp,family=quasibinomial(link="logit"), na.action="na.omit")$residuals)
      names(regcorrxp)="residuals"
      regcorrxp$gene = corrxp$gene
      regcorrxp=merge(corrxp, regcorrxp, by="gene", all.x=T)
      row.names(regcorrxp)= regcorrxp$gene
      kvals=vector()
      if(newvar != cvar && meta %in% c("PAM50", "Mammaprint_(Tian_2010)", "Protein_targets_of_FDA_appr._Drugs", "Ribosomal", "Spliceosome", "COSMIC_database_and__Nik-Zainal_93_BC_drivers", "COSMIC_BC_drivers_and_Nik-Zainal_93_BC_drivers")){
        i=i+1
        pl[[i]]=grouplogitplot(regcorrxp,newvar,"correlations", cov=c("mRNA_sd","protein_sd"), meta)+
          scale_x_continuous(sub("sd","SD",newvar))+
          scale_y_continuous("logit-transformed\nmRNA-protein correlation")
      }
      
      metacorr=regcorrxp$residuals[rownames(regcorrxp) %in% exprdata$metagene[[meta]]]
      direction = "greater"
      meanres=paste("meanres",newvar)
      pmeta[meta,meanres] = mean(metacorr, na.rm=T)
      if(pmeta[meta,meanres]<0){
        direction="less"
      }
      pmeta[meta,"direction"] = direction
      other=regcorrxp$residuals[!(rownames(regcorrxp) %in% exprdata$metagene[[meta]])]
      metacorr=wilcox.test(metacorr, other,alternative=as.vector(pmeta[meta,"direction"]))
      pmeta[meta,paste("pval",newvar)] = p.adjust(metacorr$p.value,method="BH")#,n=3*4)
      pmeta[meta,paste("fpval",newvar)] = sprintf("%.4g",metacorr$p.value)
      pmeta[meta,paste("wilcoxparam",newvar)] = metacorr$statistic
    }
  }
  pmetas[[p]]=pmeta
}  
blank <- grid.rect(gp=gpar(col="white"))

```
```{r,thefig, , fig.height=12, fig.width=7} 
#fig.cap="A-D) mRNA-protein correlation as a function of standard deviation (SD) for mRNA and protein expression values. For each statistics (see x-axis legend), A and B shows, as a function of mRNA SD, a scatterplot of untransformed correlation coefficients and a scatterplot of logit-transformed correlation with a glm-regression line, respectively, and C and D shows the same for protein SD. E-R) Distribution of selected protein groups in scatterplots of logit-transformed Spearman correlation vs mRNA or protein SD. E-F)  Mammaprint (Tian_2010), G-H) PAM50, I-J) Protein targets of_FDA appr. Drugs, K-L) Ribosomal, M-N) Spliceosome, O-P) COSMIC_database_and__Nik-Zainal_93_BC_drivers, Q-R) COSMIC_BC_drivers_and_Nik-Zainal_93_BC_drivers. Black dots indicate the specified protein group, while red dots represent all other proteins.  The black line indicates the marginal regression line for mRNA or protein sd, i.e., logit(y)~b_ix_i+c for i in (1,2), where x_1 is the mRNA SD and x_2 the protein SD, and b_1 and b_2 are the corresponding effect sizes for x_1 and x_2, respectively, estimated from the bivariate regression model logit(y)~b_1 x_1+b_2 x_2 + c."}

lay <- rbind(c(1,1,2,2),
             c(1,1,2,2),
             c(3,3,4,4),
             c(3,3,4,4),
             c(5,6,7,8),
             c(9,10, 11,12),
             c(13,14, 15,16),
             c(17,18, 19,20)
             )
grid.arrange(pl[[1]]+labs(tag="A"),pl[[2]]+labs(tag="B"),pl[[3]]+labs(tag="C"),pl[[4]]+labs(tag="D"),
             pl[[5]]+labs(tag="E"),pl[[6]]+labs(tag="F"),pl[[7]]+labs(tag="G"),pl[[8]]+labs(tag="H"),
             pl[[9]]+labs(tag="I"),pl[[10]]+labs(tag="J"), pl[[11]]+labs(tag="K"), pl[[12]]+labs(tag="L"),
             pl[[13]]+labs(tag="M"), pl[[14]]+labs(tag="N"), pl[[15]]+labs(tag="O"), pl[[16]]+labs(tag="P"),
             pl[[17]]+labs(tag="Q"), pl[[18]]+labs(tag="R"),blank, blank,layout_matrix=lay,respect=TRUE)
```

```{r,thefigpdf,results="hide"}
# pdf
outfile1=paste(out_dir,"Lehtio_S6.pdf",sep="/")
pdf(file=outfile1, height=12, width=7)
grid.arrange(pl[[1]]+labs(tag="A"),pl[[2]]+labs(tag="B"),pl[[3]]+labs(tag="C"),pl[[4]]+labs(tag="D"),
             pl[[5]]+labs(tag="E"),pl[[6]]+labs(tag="F"),pl[[7]]+labs(tag="G"),pl[[8]]+labs(tag="H"),
             pl[[9]]+labs(tag="I"),pl[[10]]+labs(tag="J"), pl[[11]]+labs(tag="K"), pl[[12]]+labs(tag="L"),
             pl[[13]]+labs(tag="M"), pl[[14]]+labs(tag="N"), pl[[15]]+labs(tag="O"), pl[[16]]+labs(tag="P"),
             pl[[17]]+labs(tag="Q"), pl[[18]]+labs(tag="R"),blank, blank,layout_matrix=lay,respect=TRUE)
dev.off()

# tiff
outfile2=paste(out_dir,"Lehtio_S6.tiff",sep="/")
tiff(file=outfile2, height=12, width=7, units="in", res=600)
grid.arrange(pl[[1]]+labs(tag="A"),pl[[2]]+labs(tag="B"),pl[[3]]+labs(tag="C"),pl[[4]]+labs(tag="D"),
             pl[[5]]+labs(tag="E"),pl[[6]]+labs(tag="F"),pl[[7]]+labs(tag="G"),pl[[8]]+labs(tag="H"),
             pl[[9]]+labs(tag="I"),pl[[10]]+labs(tag="J"), pl[[11]]+labs(tag="K"), pl[[12]]+labs(tag="L"),
             pl[[13]]+labs(tag="M"), pl[[14]]+labs(tag="N"), pl[[15]]+labs(tag="O"), pl[[16]]+labs(tag="P"),
             pl[[17]]+labs(tag="Q"), pl[[18]]+labs(tag="R"),blank, blank,layout_matrix=lay,respect=TRUE)
dev.off()
```

Saved file as pdf to `r outfile1` and as tiff to `r outfile2`



## Table S3 : Gene GSEA corr SD influence

###Legend
One-sided Wilcoxon rank sum test for difference in residual distribution between specified protein group and all other proteins for the regression with glm logit(y)~a+bx+cz+e, where y = mRNA-protein Spearman correlation and x = mRNA SD, z = protein SD, e i an error term, which is binomial distributed. Direction of test is given by the sign of the mean residual column and Bonferroni-corrected significance level α=0.00192 (26 tests).

The test is used in two independent analyses:

A) Pre-defined protein groups

B) Protein groups from enrichment that display significant enrichment for high RNA-protein correlation.

### The table
```{r, tabv,eval=TRUE}
myvars=list(paste0(c("mRNA_sd","protein_sd"),collapse="+"),"mRNA_sd","protein_sd")
meta4output = pmetas[["Predefined"]][, c("name","meancorr",apply(expand.grid(c("fpval","meanres"),myvars), 1, function(x) paste(x[1],x[2], sep=" ")))]
names(meta4output) = c("Protein group","Mean Spearman correlation",apply(expand.grid(c("Pvalue","Mean residual logit-transformed correlation"),gsub("sd","SD",myvars)), 1, function(x) paste(x[2],x[1], sep="                                      ")))
kable(meta4output, align="l",row.names=FALSE, caption = "A) Pre-defined proteing groups")

meta4output = pmetas[["Enriched"]][, c("name","meancorr",apply(expand.grid(c("fpval","meanres"),myvars), 1, function(x) paste(x[1],x[2], sep=" ")))]
names(meta4output) = c("Protein group","Mean Spearman correlation",apply(expand.grid(c("Pvalue","Mean residual logit-transformed correlation"),gsub("sd","SD",myvars)), 1, function(x) paste(x[2],x[1], sep="                                      ")))
kable(meta4output, align="l",row.names=FALSE, caption="B) Protein groups from enrichments")
```

Save Table 'Predefined' to  "Lehtio_vS6_Predefined.XLSX" and Table 'Enriched' to "Lehtio_S6_Enriched.XLSX".

```{r, writevtable,results="hide"}
p="Predefined"
outfile=paste(out_dir,paste("Lehtio_S3_",p,".XLSX",sep=""),sep="/")
meta4outputpredefined = pmetas[[p]][, c("name","meancorr",apply(expand.grid(c("fpval","meanres"),myvars), 1, function(x) paste(x[1],x[2], sep=" ")))]
names(meta4outputpredefined) = c("Protein group","Mean Spearman correlation",apply(expand.grid(c("Pvalue","Mean residual logit-transformed correlation"),gsub("sd","SD",myvars)), 1, function(x) paste(x[2],x[1])))
WriteXLS(meta4outputpredefined, ExcelFileName=outfile,)

p="Enriched"
outfile=paste(out_dir,paste("Lehtio_S3_",p,".XLSX",sep=""),sep="/")
meta4outputenriched = pmetas[[p]][, c("name","meancorr",apply(expand.grid(c("fpval","meanres"),myvars), 1, function(x) paste(x[1],x[2], sep=" ")))]
names(meta4outputenriched) = c("Protein group","Mean Spearman correlation",apply(expand.grid(c("Pvalue","Mean residual logit-transformed correlation"),gsub("sd","SD",myvars)), 1, function(x) paste(x[2],x[1])))
WriteXLS(meta4outputenriched, ExcelFileName=outfile,)
```







## Supplementary plot S12
Distribution of mRNA-protein Spearman correlation over tumours.For each tumour, the correlations are computed across genes:
A) Correlations based on all genes
B) Correlations based on only the 300 genes with showing the highest mRNA-protein correlation across tumours.
Tumours are ordered according to their mRNA-protein correlation and are colour-labeled by their PAM50 classification. The horizontal bar shows the mean mRNA-protein correlation over tumours.

```{r, figviii, eval=TRUE}

# for TUMOURS
# tumours ordered by correlation value, and labeled by PAM50 type 
p1=tumourplot(corr_tumour,x="tumour", y="correlations", fill="PAM50")+
  scale_x_discrete("Tumour sample")+
  scale_y_continuous("mRNA-protein correlation")
  

# tumours ordered by correlation value, and labeled by PAM50 type 
p2=tumourplot(corr_tumour,x="tumour", y="Best300.correlations", fill="PAM50")+
  scale_x_discrete("Tumour sample")+
  scale_y_continuous("mRNA-protein correlation")

lay <- rbind(c(1,2))
grid.arrange(p1+ggtitle("A"),p2+ggtitle("B"),layout_matrix=lay,respect=TRUE)
```


```{r,figviiipdf,results="hide"}
outfile="Lehtio_S12.pdf"
pdf(file=paste(out_dir,outfile,sep="/"))
grid.arrange(p1+ggtitle("A"),p2+ggtitle("B"),layout_matrix=lay,respect=TRUE)
dev.off()
```

Saved fig as `r outfile`



## Main figure 5

###Legend:
(A) Correlation between protein and mRNA quantitative values (Spearman) of individual genes. The mean correlation is indicated by a vertical line. Grey fill color indicate significanct mRNA-protein correlation (Bonferroni-Holm-adjusted P-values).

(B) Distribution of mRNA-protein correlations for selected groups of genes. Mean values are indicated by a horizontal bar, boxes indicate the 0.25 and 0.75 quantiles, and vertical bars indicate values within 1.5 * interquartile range; values outside this range are indicated by dots. P-values at top indicate significant enrichment ($alpha=0.00192;Bonferroni-adjusted for 26 test; direction is given by sign of the mean); for additional gene groups and mRNA-protein correlation analysis considering data distribution, see Fig S10, 11 and Table S3. 
Direction of test is given by the sign of the mean residual column and Bonferroni-corrected significance level α=0.00192 (26 tests).

###Notes:
1) cutoff p<=0.05 does not map exactly between two bars. 
2) Notice that there are also significant negative correlations

```{r fig5}
#A
# genes ordered by correlation value, and labeled by FDR
signlevel = 0.05
myalpha=1
# Determine cuoff for correlation value representing significance
p_corr_cutoff = min(abs(corr_genes[corr_genes$Padjust < signlevel, "correlations"]))
p1=ggplot(corr_genes,aes(x=correlations))+
  scale_fill_grey(start=0,end=1)+
  geom_histogram(aes(alpha=myalpha,fill=(abs(correlations)<=p_corr_cutoff) ,y = 100*..count../sum(..count..)), bins=30, color="black")+
  scale_x_continuous(expand = c(0, 0),"mRNA vs protein correlation")+
  scale_y_continuous(expand = c(0, 0),"Frequency (%)", limits=c(0,9))+
  theme_classic(base_size=8)+
  theme(axis.line.y = element_line(colour = "black", linetype="solid", size=0.5),axis.line.x = element_line(colour = "black", linetype="solid", size=0.5))+
  theme(plot.margin=unit(c(0,0,0.7,0),"cm"))+
  guides(fill=FALSE)+
  geom_vline(aes(xintercept=mean(corr_genes$correlations)))+
  ggtitle("A")+
  theme(plot.title = element_text(hjust=-0.1,margin = margin(t = 0, b = -10)))+ 
  theme(legend.justification=c(0,1),legend.position=c(0,1),legend.title=element_blank(), legend.key.size=unit(5, "pt"))+
  scale_alpha_continuous(breaks=c(1),labels=c("Significant correlation p<0.05"))+
  annotate("text",x=-0.45,y=7.4,hjust=0,vjust=1,label=paste("n=",length(corr_genes$correlation),"\nmedian=",sprintf("%.2f",median(corr_genes$correlation)),"\nmean=",sprintf("%.2f",mean(corr_genes$correlation)),sep=""), size=2)



#B
sel_metagroups=data.frame(alias=c(
                    "Mammaprint",
                    "Oncotype DX",
                    "PAM50",
                    "Ribosome",
                    "Spliceosome",
                    "Transcription factors",
                    "Transporters",
                    "Targets of FDA appr. Drugs"),
                    row.names=c(
                      "Mammaprint_(Tian_2010)",
                    "Oncotype_DX",
                    "PAM50",
                    "Ribosomal",
                    "Spliceosome",
                    "Transcription_factors",
                    "Transporters",
                    "Protein_targets_of_FDA_appr._Drugs")
)

groupcorr = data.frame(group="All", value=corr_genes$correlations)
coord=1.0
# pvals=data.frame(group="All", value=coord, p=1)
pvals=data.frame(group="All",p="p-value")
for(g in row.names(sel_metagroups)){
  tmpgenes=intersect(genes,unlist(exprdata$metagene[[g]]))
  #g1=gsub("_"," ",g)
  g1=sel_metagroups[g,]
  tmp=data.frame(group=g1, value=corr_genes$correlations[rownames(corr_genes) %in% tmpgenes])
  groupcorr=rbind(groupcorr, tmp)
  p=pmetas[["Predefined"]][g,"fpval mRNA_sd+protein_sd"]
#  p=wilcox.test(corr_genes$correlations[rownames(corr_genes) %in% tmpgenes],corr_genes$correlations)$p.value
  if(as.numeric(p)>0.00625){
    p=NA
  }
  pvals=rbind(pvals, data.frame(group=g1,p=p))
}

# fix the order of factor group
groupcorr$group = factor(groupcorr$group, c("All",
                    "Mammaprint",
                    "Oncotype DX",
                    "PAM50",
                    "Ribosome",
                    "Spliceosome",
                    "Transcription factors",
                    "Transporters",
                    "Targets of FDA appr. Drugs"), ordered=T)
#plot them
p2=bxplot(groupcorr, group, value)+
  scale_x_discrete(element_blank())+
  scale_y_continuous("mRNA vs protein correlation")+
  theme(plot.margin=unit(c(0,0,1.2,0),"cm"))+
  theme(plot.title = element_text(hjust=-0.1,margin = margin(t = 0, b = -5)))+ggtitle("B")+
  geom_text(data = pvals, aes(x=group,y=1.0, label=p), size=1.5, angle=0)+
  guides(fill=FALSE)

lay <- rbind(c(1,2),
             c(3,2)
             )
blank <- grid.rect(gp=gpar(col="white"))
grid.arrange(p1,p2,blank, layout_matrix = lay, heights=unit(c(6,2.3),"cm"))

```


```{r,fig5pdf,results="hide"}
outfile="Lehtio_5.pdf"
pdf(file=paste(out_dir,outfile,sep="/"))
grid.arrange(p1,p2,blank, layout_matrix = lay, heights=unit(c(6,2.3),"cm"))
dev.off()
```

Saved fig as `r outfile`